Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-02-25 tpe_edreview_rebuttal.txt

I was unaware of the derivation in Taylor and Wheeler that the
reviewer points me to.  I did not want to suggest that equation 2 was
new, but was unable to find a derivation of it in the literature.  I
have cited Taylor and Wheeler (but see my response to Q6 below).

> 
> The paper raises a few questions in this reviewer's mind, and in
> elsewhere it offers opportunities for suggestions.
> 
> 1.  Question.  First paragraph in "Introduction": The third sentence
> "Here I interpret straight line as the path adopted by a stationary,
> light inextensible string under tension passing between two points."
> Is this to be understood as identical to the definition of "geodesic"?
> If that is the assumption, can it be justified, because when one
> thinks of geodesic in General Relativity, one has "free-fall" in mind,
> and the stretched string, being "stationary," would not be in
> free-fall in the global coordinate system.
> 
> 2.  Suggestion about the line that begins on p. 1 and continues to
> p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
> present a sticking point for the undergraduate audience towards whom
> this paper in intended.  The logic behind this inequality would show
> greater clarity if one says something like "the weight of the string
> does not exceed the tension, so that $L\mu g<T$".  From this the
> stated inequality about lengthscales follows easily.
>

This is a much better way of phrasing the argument and I have
rephrased as suggested.

> 3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
> event horizon" may leave an undergraduate audience baffled.  By
> starting from where they are, with $g=GM/r^2$, then setting r equal to
> the Schwarzschild radius, the stated result follows immediately,
> without mystery.
>

I have totally rephrased that part of the manuscript using this
suggestion and it is much clearer (I have also changed uppercase L for
lowercase l as per suggestion number 7.

> 4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
> by "the dominant energy condition"?

On reflection the dominant energy condition is not really an
appropriate or informative or useful or interesting concept for the
intended readership.  The important thing is that tension greater $\mu
c^2$ is unphysical and this context, the unphysicality is manifested
by transverse waves travelling faster than light.

In answer to the question, we may take the stress-energy tensor for a
stretched string to be diagonal with elements $\mu,-T,0,0$.  Then the
dominant energy condition is that, for any future-pointing causal
vector, $-T^\mu_\nu V^\nu$ must also be a future-pointing causal
vector.  We can see directly that if $T>\mu$ the DEC is violated.

But this is a massive digression to the main thrust of the manuscript
and any adequate pedagogical explanation would be an intrusive
distraction.  It is like using a slegehammer to crack a nut.  I have
removed any reference to the DEC on these grounds, and further
emphasised the unphysicality of the wave speed being $>c$.

> 
> 5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
> circular string at constant r will, trivially, have proper length
> $2\pi r$."  Stating the relation this way muddies the logic and the
> distinction between flat and Schwarzschild space.  The r-coordinate of
> a spherical shell centered on a point mass M is defined operationally
> by measuring the circumference of the shell and dividing the
> circumference's value by $2\pi$.  That must be understood before the
> calculations of Eq. (2) can make sense.  As the sentence is presently
> structured it may obscure the distinction between the r coordinate and
> radial lengths because it invites the undergraduate reader to
> interpret the "triviality" as a statement about plane Euclidean
> geometry instead of the Schwarzschild geometry's definition of the
> r-coordinate.
> 
> 6.  Question.  There may be a typographical error in Eq. (2).
> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
> written the equation is not dimensionally consistent, since r and M
> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
>

### 2021-02-25 tpe_edreview_rebuttal.txt

I was unaware of the derivation in Taylor and Wheeler that the
reviewer points me to.  I did not want to suggest that equation 2 was
new, but was unable to find a derivation of it in the literature.  I
have cited Taylor and Wheeler (but see my response to Q6 below).

> 
> The paper raises a few questions in this reviewer's mind, and in
> elsewhere it offers opportunities for suggestions.
> 
> 1.  Question.  First paragraph in "Introduction": The third sentence
> "Here I interpret straight line as the path adopted by a stationary,
> light inextensible string under tension passing between two points."
> Is this to be understood as identical to the definition of "geodesic"?
> If that is the assumption, can it be justified, because when one
> thinks of geodesic in General Relativity, one has "free-fall" in mind,
> and the stretched string, being "stationary," would not be in
> free-fall in the global coordinate system.

I think the sentence in question was technically correct but
potentially misleading.  What I wanted to do was to make readers think
about taut thread in the context of GR.  I didn't want to consider
geodesics at this point in the introduction, or free-falling elevator
frames.  I want to focus on taut string.

[one of the conceptual difficulties in preparing the manuscript, as
pointed out by a previous reviewer, is that the string is not in a
freely-falling elevator frame; as the reviewer points out, such string
does not follow a geodesic path in any meaningful sense].

I have rephrased the sentence which now reads: "Here I consider the
path adopted by a stationary, light inexensible string under tension
passing between two points, and consider the extent to which the
expression "straight line" would be applicable to such an object".

> 
> 2.  Suggestion about the line that begins on p. 1 and continues to
> p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
> present a sticking point for the undergraduate audience towards whom
> this paper in intended.  The logic behind this inequality would show
> greater clarity if one says something like "the weight of the string
> does not exceed the tension, so that $L\mu g<T$".  From this the
> stated inequality about lengthscales follows easily.
>

### 2021-02-25 string_physics_educator_revision3.tex
fixes #37
fixes #34

General relativity is a topic of long-standing fascination for physics
undergraduates~\cite{christensen2012,taylor2000}.  Students appreciate
everyday illustrations of the concepts involved, as recent educational
work on gravitational wave visualisations attests~\cite{overduin2018}.
Here I interpret a straight line as the path adopted by a stationary,
light inexensible string under tension passing between two points.
The analysis presented here would be suitable for an undergraduate
homework problem~\cite{romano2019}.

@@

Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string of length $\ell$ will behave
as though light if the weight of the string does not exceed the
tension, so that $T\gtrsim\ell\mu g$.  At the Schwarzschild radius
$2GM/c^2$ of a black hole with mass $M$, the Newtonian value
$g=GM/r^2$ gives $T\gtrsim\ell\mu c^4/M$.  For the string to behave as
though light on lengthscales comparable with the Schwarzschild radius
(as in the simulations below), we would require $T\gtrsim\mu c^2$.
Exceeding this criterion would violate the dominant energy condition
(and in this context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: Rindler~\cite{rindler}, for example, considers lowering a heavy
object to the event horizon on the end of such a string to illustrate
energetic arguments in GR.

>>>>

General relativity is a topic of long-standing fascination for physics
undergraduates~\cite{christensen2012,taylor2000}.  Students appreciate
everyday illustrations of the concepts involved, as recent educational
work on gravitational wave visualisations attests~\cite{overduin2018}.
Here I consider the path adopted by a stationary,
light inexensible string under tension passing between two points, and consider the extent to which the expression  ``straight line'' would be applicable to such an object.
The analysis presented here would be suitable for an undergraduate
homework problem~\cite{romano2019}.

@@

Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string of length $\ell$ will behave
as though light if the weight of the string does not exceed the
tension, so that $T\gtrsim\ell\mu g$.  At the Schwarzschild radius
$2GM/c^2$ of a black hole with mass $M$, the Newtonian value
$g=GM/r^2$ gives $T\gtrsim\ell\mu c^4/M$.  For the string to behave as
though light on lengthscales comparable with the Schwarzschild radius
(as in the simulations below), we would require $T\gtrsim\mu c^2$.
Exceeding this criterion is unphysical:
transverse waves propagate at
speed~$\sqrt{T/\mu}$ which would exceed $c$.  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: Rindler~\cite{rindler}, for example, considers lowering a heavy
object to the event horizon on the end of such a string to illustrate
energetic arguments in GR.
